
# Stratified Analysis of 2x2 Tables: Practical 13

In this practical we look at stratified analysis of 2x2 tables to analyse the association between two binary variables in the presence of a confounding variable.

As always, we commence by loading the appropriate libraries, changing any options, and reading in the dataset.

```{r, results = FALSE, warning = FALSE, message = FALSE}
#--- Load libraries
library(epiDisplay)
library(foreign)
library(psych)
library(tidyverse)

#--- Change to show no scientific notation & round to 3 decimal places
options(scipen = 10, digits=3) 

#--- Set the plot theme to come out in black and white
theme_set(theme_bw())

#--- Read in the file
creche <- read.dta("./CRECHE.dta", convert.factors = T) 
```

We wish to investigate whether attending a creche is associated with having had a lower respiratory tract infection.

This dataset contains socioeconomic status information in the variable 'socio', information on whether an individual was cared for at home (no) or attends a creche (Yes) in the variable 'creche' and whether in the past year an individual had had a respiratory infetion in the variable 'lrti'. Note that these variables are coded as factors - R's way of coding categorical variables.

The key commands here are cc() from EpiDisplay and tabyl() from janitor, both of which will enable you to get a 2x2 table. The cc() function also conducts some hypothesis tests and extracts an OR. To assess the strength of evidence for an association between lower respiratory tract infection and creche attendance, we can look at the result of the $\chi^2$ test. 

```{r}
#--- Get a 2x2 table with percentages
tabpct(creche$lrti, creche$creche, graph = F)

#--- Get a 2x2 table with totals, tests and an OR
cc(creche$lrti, creche$creche, graph = F)

#--- Using the magrittr package, we can write this more neatly using the %$% pipe operator
#--- This pipe operator takes the object on the left and adds it in front of the objects on the right
#--- with a dollar sign in between and generally results in more readable code
creche %$% tabpct(lrti, creche, graph = F)
creche %$% cc(lrti, creche, graph = F)
```

```{block2, type='rmdexercise2'}
Exercise 13.1: Is there an association between creche attendance and lower respiratory tract infection?
```

Lower respiratory tract infection is strongly related to socioeconomic status (SES). If creche attendance is also related to SES we might expect confounding from SES to impact the association between lower respiratory tract infection and socioeconomic status. 

To handle this confounding, we can analyse the association separately within strata defined by the confounder (stratification). 

```{block2, type='rmdexercise2'}
Exercise 13.2: Use the cc() command to determine if a) creche attendance is associated with SES and b) lower respiratory tract infection is associated with SES. What do you conclude about confounding?
```

We can examine the 2x2 tables by group by filtering the dataset to the stratum we wish to investigate and then using the tabpct command.

```{r}
creche %>% filter(socio == "Average SES") %$% tabpct(lrti, creche, graph = F)
creche %>% filter(socio == "Very low SES") %$% tabpct(lrti, creche, graph = F)
```

```{block2, type='rmdexercise2'}
Exercise 13.3: What percentage of children attending a creche had lower respiratory tract infection if they were from a higher SES? A lower SES?
```

While the cc() command calculates the odds ratio for you, it is useful to note that in R you can simply do this 'by hand'. There is no need to use the display command like in Stata. 

```{r}
#--- Get the 2x2 table
creche %$% cc(lrti, creche, graph = F)

#--- Calculate the odds and odds ratio 'by hand'
27/62 # odds of LRTI in creche goers
53/499 # odds of LRTI in creche goers
(27/62)/(53/499) # OR
(499*27)/(62*53) # Alternate calculation of OR

```

We saw that socio-economic status may be confounding the relationship between creche attendance and lower respiratory tract infection. Thus we should examine the odds ratio for the effect of creche attendance on lower respiratory tract infection in each stratum of SES separately. We can then combine these estimates by taking a weighted average to come up with a summary value. We call this the effect of creche attendance controlled for socioeconomic status - in the case of odds ratios, we call this the Mantel-Haenszel odds ratio.

In R, we use epiDisplay's mhor command to conduct this test, providing the odds ratio, 95% confidence interval, and p-value for each stratum and overall. The test for homogeneity of ORs tests the hypothesis that the ORs in the strata are the same (i.e. it tests for effect modification)

```{r}
#--- Get the Mantel-Haenszel OR
creche %$% mhor(lrti, creche, socio, graph = F)
```

```{block2, type='rmdexercise2'}
Exercise 13.4: Using the results above: a) what is the MH odds ratio controlling for SES and what does it mean? b) How does this differ from the OR for the effect of creche attendance on lower respiratory tract infection ignoring SES? c) Why has the OR decreased? d) Interpret the 95% CI for the MH OR.
```

Sometimes, we may wish to use data from prepared tables instead of individual rows. Let's examine the use of bednets and the presence of an enlarged spleen. We want to manually enter four columns and we can do this using the command c() -- 'c' stands for concatenate. 

```{r}
village <- c(1,1,1,1,2,2,2,2)
spleen <- c(1,1,0,0,1,1,0,0)
bednet <- c(1,0,1,0,1,0,1,0)
freq <- c(12,42,12,29,15,4,52,12)

```

