
# Visualising Data

This section will take you through some basic data visualisation in R using the ggplot2 package.

## Setup & Reading in Data

We are going to examine a number of different questions surrounding the Sustainable Development Goals in this session.

The first thing we are going to do is make sure we have the packages we need installed and the libraries loaded. So we set up a chunk of code to ensure the tidyverse is loaded.

```{r}
#--- Load libraries
library(tidyverse)
```

To start investigating our question, we need a dataset. Most data comes to you in a csv (comma separated values) file from Excel - we will talk about other formats later. This dataset, sdg.csv, should've come with the downloaded zip file - make sure that it is located in the same folder as this R file.

We use a simple function to read in data: read.csv(). 


```{r}
#--- Read in the data
sdg <- read.csv("./sdg.csv")

#--- This will only work on my computer
# sdg <- read.csv("E:/LSHTM/Teaching/r4epi/sdg.csv")

#--- Get working directory
getwd()
```

Let's unpack a couple of things from this code. A key idea in R is that objects (like a dataset) are assigned to particular names. We read in the data and assigned it the name 'sdg' because it is a collection of data on a variety of Sustainable Development Goal subtargets. We did the assignment using the '<-' arrow. The function read.csv() took one argument, where we told R which file to read in. 

Note that we did not specify the full path to the data. We use the "." to tell R to look in the same directory where the R file you are working in is saved. We can view what directory this is with getwd() -- wd stands for "working directory". In general, you should never have to change the working directory explicitly - just make sure you save your data in the same place as your analysis file.

When you read in the data, you should've seen a new object in the top right Environment pane. 

> EXERCISE: How many rows does the dataset have? What does each row represent? How many columns does the dataset have? What does each column represent?

```{r}

```


## Reading in a dta file

You might also be provided with data in a .dta format if someone has exported it from Stata for you. There are some slight differences in reading in dta files compared to csv files. Let's read in a dta file now using the readstata13 package.

> Exercise: Install and load the foreign package (go back to getting_started_STEPH if you need to)

```{r}

```

We can read in a dataset from Stata using read.dta(). Where possible, setting convert.factors = T should read in Stata labels for factors. Remember that you will need to change the path to where the data is stored on your computer. If it is in the same folder as your R file, you can use the . abbreviation.

```{r}
#--- Read in a Stata file
bab9 <- read.dta("E:/LSHTM/Teaching/r4epi/TEMP/BAB9.DTA", convert.factors = T)
```

Simple! Just like reading in a csv file. For Stata 13 files, read.dta13() is useful, and you will need to set the option nonint.factors = T. Note that we now have two datasets stored in the Environment (top right pane).

## Initial Investigations

We might want to look at the SDG data. We can do that by clicking on the name of the dataset in the Environment pane, or with the following code:

```{r}
#--- Look at the dataset (note the capitalisation, R is case sensitive!)
View(sdg)

#--- We can also just look at the first few rows (useful for large datasets)
head(sdg)

#--- We can also just get the column names
names(sdg)

```

Note the well-commented code telling you what each snippet does!

Since we are interested in TB, we might want to look at some TB-specific data. The column 'tb' contains information about the TB incidence rate expressed as the number of cases per 100,000 people.

```{r}
#--- Look at the TB values for the first few observations
head(sdg$tb)

#--- Get some summary statistics for the values of TB incidence
summary(sdg$tb)
```

Note that here we use a dollar sign to "index" the 'tb' column. Typing "sdg$tb" says that we are interested in the column tb in the dataset sdg. 

> EXERCISE: What is the mean GDP for all nations? How many nations have missing GDP data?

```{r}

```


We can get some more detailed summary information using the psych package. 

> EXERCISE: Install the psych package and load its library

```{r}

```


```{r}
#--- Get a detailed summary
describe(sdg$tb)

#--- Get a summary by group
describeBy(sdg$tb, sdg$lmic)
```

> EXERCISE: Get detailed summary statistics for maternal mortality.

## A Pipe Aside

If you were typing out this code, you'd be very tired of typing the dataset name by now. Enter the %$% pipe from the magrittr package. Remember how we said we access columns from a dataset using the dollar sign? This pipe puts whatever is on the left of the pipe in front of whatever is on the right of the pipe, with a dollar sign in between.

```{r, echo = F, warnings = F}
#--- Load the magrittr library - remember to install it first if you need to
library(magrittr)
```

Convince yourself that the pipe does this with the below code. You don't have to use this pipe, but it can make using tidyverse functions and base R together much easier, and gives you a convenient syntax for always starting with the dataset as the leftmost argument.

If you want to know more about the describe function, you can get help by typing ?describe into the Console. This is true for any function in R.

```{r}
#--- Get a detailed summary
describe(sdg$tb)

#--- Same thing, but with a pipe
sdg %$% describe(tb)

#--- Get the correlation between TB and Gini coefficient
## Note that the 'use' argument says to only use data where both observations are not missing
cor(sdg$tb, sdg$gini, use = "complete.obs")

## This correlation says that TB & ARI are slightly positively associated.
## As income inequality goes up, so does TB (this shouldn't be surprising!)

#--- The same code with a pipe
sdg %$% cor(tb, gini, use = "complete.obs")

```


## Starting to Plot

Let's explore further the relationship between the Gini index with TB incidence. Let's explore this visually. Here we are going to introduce the ggplot2() package. 

In R, we will often build our analyses step by step, appending small chunks of code together to complete larger tasks. It is the same concept with visualisation. Consider this like painting: we combine elements like line and shape and color together into a cohesive whole. 

Let's start with a very basic plot:

```{r}
#--- Our first scatterplot
ggplot(data = sdg, aes(x = gini, y = tb)) +
  geom_point()
```

Let's unpack the code here. The ggplot() function is the one that builds the plot. We tell it that we want to use the sdg dataset we loaded in. We also need to tell ggplot how to map the various attributes of the plot to attributes of the data. We do that mapping with the aes() command, standing for aesthetics. We tell ggplot() that on our x axis we want the Gini coefficient and on the Y axis we want TB incidence. Note the handy warning that we've got missing data that hasn't been plotted!

Now we have to tell ggplot what type of graph we want -- we do this with geom_point(), which tells ggplot to produce a scatterplot. 

Let's say we want to look at country aggregate GDP using the lmic variable. Count returns a frequency table. We can also visualise with a bar chart.

```{r}
#--- Get frequency tables and bar chart
sdg %>% 
  count(lmic)

ggplot(sdg, aes(lmic)) +
  geom_bar() +
  theme_bw()
```

## A bit more detail...

We might want to make our scatterplot a bit prettier and convey some more information, so we add a new mapping: we tell ggplot to color the points by the WHO region of interest. To make our points clear, we change the 'theme' of our plot using theme_bw() -- there are several default plot themes inbuilt into R.

```{r}
#--- Colour the points and remove the background grey
ggplot(data = sdg, aes(x = gini, y = tb, color = reg)) +
  geom_point() +
  theme_bw()
```



## ...Much more detail

What if we wanted to convey even more information from this plot? Let's add the population of each country as a size element. We'll also give our plot a title and change the labels. 

```{r}
#--- Add titles and add size aesthetic
ggplot(data = sdg, aes(x = gini, y = tb, color = reg, size = pop)) +
  geom_point() +
  theme_bw() +
  labs(title = "Income Inequality vs TB Incidence",
       x = "Gini Coefficient",
       y = "TB Incidence (cases per 100,000)",
       color = "WHO Region",
       size = "Population")
```

> EXERCISE: Create a similar plot to the above, but looking at the relationship between TB and the proportion who live in slums. Give the variable coding categorised GDP (low/medium/high income countries; lmic) as the color and the TB case detection rate as the size

```{r}

```


> EXERCISE: Modify the same plot to show categorised GDP with a shape parameter instead of a color parameter. Label the plot.

```{r}

```


> CHALLENGE EXERCISE: Add a smoothed line to the plot to display the relationship between TB and the proportion who live in slums. 

```{r}

```


